---
title: "A Practical Workflow for Dose-Response Analysis"
author: "Hannes Reinwald"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteTitle{A Practical Workflow for Dose-Response Analysis}
  %\VignetteIndexEntry{A Practical Workflow for Dose-Response Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(drc)
```

## Executive Summary

This vignette provides a comprehensive, step-by-step workflow for conducting proper dose-response analysis using the `drc` package. We demonstrate the complete analysis process from initial model fitting through model selection, validation, and interpretation. By following this workflow, even inexperienced users can perform rigorous dose-response modeling while avoiding common pitfalls.

## Introduction

Dose-response analysis is fundamental in toxicology, ecotoxicology, pharmacology, and related fields. The relationship between dose (or concentration) and biological response often follows non-linear patterns that require specialized statistical models. The `drc` package provides a comprehensive framework for fitting, comparing, and interpreting dose-response models.

### What You Will Learn

This vignette demonstrates a complete workflow including:

1. Initial exploratory model fitting
2. Visual assessment of model adequacy
3. Statistical evaluation of model fit
4. Systematic model comparison and selection
5. Model-averaged estimation for robust inference
6. Understanding the impact of parameter constraints
7. Choosing appropriate models for different data types

### The Example Dataset

We will use the `ryegrass` dataset, which contains measurements of root length in perennial ryegrass (*Lolium perenne L.*) exposed to different concentrations of ferulic acid, a phenolic compound that inhibits plant growth.

```{r load-data, fig.alt="Scatter plot showing ryegrass root length (cm) versus ferulic acid concentration (mM), displaying a decreasing dose-response relationship"}
# Load the ryegrass dataset
data(ryegrass)

# Examine the data structure
head(ryegrass, 10)

# Summary statistics
summary(ryegrass)

# Simple exploratory plot
plot(rootl ~ conc, data = ryegrass,
     xlab = "Ferulic acid concentration (mM)",
     ylab = "Root length (cm)",
     main = "Ryegrass Root Growth vs. Ferulic Acid Concentration",
     pch = 16, cex = 1.2)
```

The dataset contains 24 observations with:
- `conc`: Ferulic acid concentration in millimolar (mM)
- `rootl`: Root length in centimeters (cm)

We observe a clear dose-response relationship: as the concentration increases, root length decreases, indicating an inhibitory effect of ferulic acid on ryegrass root growth.

## Step 1: Initial Model Fitting

### Choosing a Starting Model

For a typical monotonic dose-response curve, the four-parameter log-logistic model (`LL.4`) is an excellent starting point. It is flexible, well-characterized, and commonly used in toxicology.

The `LL.4` model has the form:

$$f(x) = c + \frac{d-c}{1 + \exp(b(\log(x) - \log(e)))}$$

where:
- **b**: Slope parameter (steepness of the curve)
- **c**: Lower asymptote (response at infinite dose)
- **d**: Upper asymptote (response at zero dose, control response)
- **e**: ED50 or EC50 (dose producing 50% of the maximal effect)

### Fitting the Initial Model

```{r initial-fit}
# Fit a four-parameter log-logistic model
ryegrass.LL4 <- drm(rootl ~ conc, data = ryegrass, fct = LL.4())

# Display model summary
summary(ryegrass.LL4)
```

The summary provides:
- Parameter estimates and their standard errors
- Residual standard error
- Model convergence information

**Interpretation of Parameters:**
- The `d` parameter (upper limit) represents the control root length (at zero concentration)
- The `c` parameter (lower limit) represents the minimum root length at high concentrations
- The `e` parameter (ED50) is the concentration causing 50% reduction from control
- The `b` parameter controls the steepness of the dose-response curve

## Step 2: Visual Assessment of Model Fit

Visual diagnostics are crucial for assessing whether the fitted model adequately describes the data. We use two primary tools: the standard dose-response plot and quantile-quantile (Q-Q) plots.

### Standard Dose-Response Plot

```{r basic-plot, fig.alt="Dose-response curve showing LL.4 model fit to ryegrass data with observed data points and fitted sigmoid curve"}
# Plot the fitted model with data points
plot(ryegrass.LL4, type = "all",
     main = "LL.4 Model Fit to Ryegrass Data",
     xlab = "Ferulic acid concentration (mM)",
     ylab = "Root length (cm)",
     lwd = 2, cex = 1.2)
```

The plot shows:
- Observed data points
- Fitted dose-response curve
- Overall pattern of fit

**What to Look For:**
- Do the fitted values follow the general trend of the data?
- Are there systematic deviations (e.g., all points above or below the curve in certain regions)?
- Are there outliers that might influence the fit?

### Quantile-Quantile (Q-Q) Plot

Q-Q plots assess whether the model residuals follow a normal distribution, which is an assumption of the fitting procedure.

```{r qq-plot, fig.alt="Normal Q-Q plot of residuals from LL.4 model showing points approximately along the diagonal reference line"}
# Create Q-Q plot for residual diagnostics
qqnorm(residuals(ryegrass.LL4),
       main = "Normal Q-Q Plot of Residuals (LL.4)",
       pch = 16, cex = 1.2)
qqline(residuals(ryegrass.LL4), col = "red", lwd = 2)
```

**Interpretation:**
- Points should fall approximately along the diagonal line
- Systematic deviations suggest non-normality of residuals
- Deviations at the extremes are common and often acceptable
- Severe deviations may indicate model inadequacy or outliers

### Residual Plot

An additional useful diagnostic is plotting residuals against fitted values:

```{r residual-plot, fig.alt="Residual plot showing residuals versus fitted values with random scatter around zero horizontal line"}
# Residuals vs. Fitted values
plot(fitted(ryegrass.LL4), residuals(ryegrass.LL4),
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residual Plot (LL.4)",
     pch = 16, cex = 1.2)
abline(h = 0, col = "red", lwd = 2, lty = 2)
```

**What to Look For:**
- Random scatter around zero (no systematic pattern)
- Constant variance across fitted values (homoscedasticity)
- No obvious outliers or influential points

## Step 3: Statistical Evaluation of Model Fit

Beyond visual assessment, we use formal statistical tests to evaluate model adequacy and significance.

### Test for Dose Effect: noEffect()

The `noEffect()` function performs a likelihood ratio test comparing the dose-response model to a null model (no dose effect).

```{r no-effect}
# Test whether there is a significant dose effect
noEffect(ryegrass.LL4)
```

**Interpretation:**
- The null hypothesis is "no dose effect" (all responses are equal)
- A significant p-value (< 0.05) indicates that the dose-response model fits significantly better than the null model
- This confirms that ferulic acid concentration has a significant effect on root length

### Goodness-of-Fit Test: modelFit()

The `modelFit()` function assesses whether the model adequately describes the data using a lack-of-fit test.

```{r model-fit}
# Perform goodness-of-fit test
modelFit(ryegrass.LL4)
```

**Interpretation:**
- This test compares the fitted model to a saturated model (perfect fit)
- A **non-significant** p-value suggests adequate fit (model is not significantly worse than perfect fit)
- A significant p-value indicates lack of fit (model may be inadequate)
- **Note:** This test requires replication at dose levels

### Estimating Effective Doses: ED()

Effective dose (ED) or effective concentration (EC) values are key outputs in dose-response analysis. They represent the dose required to produce a specified level of effect.

```{r ed-estimates}
# Estimate EC10, EC20, and EC50 with 95% confidence intervals
# Using delta method for confidence intervals
ed_values <- ED(ryegrass.LL4, respLev = c(10, 20, 50), interval = "delta")
ed_values
```

**Understanding ED Values:**
- **EC10**: Concentration causing 10% effect (reduction in root length)
- **EC20**: Concentration causing 20% effect
- **EC50**: Concentration causing 50% effect (often used as a summary measure of potency)

**Confidence Intervals:**
- The `interval = "delta"` argument uses the delta method for CI estimation
- Alternative methods include `"fls"` (fieller), `"tfls"` (transformed fieller)
- Narrower CIs indicate more precise estimates

### Alternative Confidence Interval Methods

```{r ed-comparison}
# Compare different confidence interval methods
cat("Delta method:\n")
ED(ryegrass.LL4, respLev = 50, interval = "delta")

cat("\nFieller method:\n")
ED(ryegrass.LL4, respLev = 50, interval = "fls")
```

The Fieller method is often preferred for ED50 estimation as it accounts for the ratio nature of the parameter.

## Step 4: Model Comparison and Selection

A critical step in dose-response analysis is comparing alternative models to select the most appropriate one. Different model families may fit the data better depending on the underlying biological mechanism.

### Comparing Multiple Models

We'll compare the initial LL.4 model with several alternatives:

- **LN.4**: Four-parameter log-normal model
- **W1.4**: Four-parameter Weibull type 1 model
- **W2.4**: Four-parameter Weibull type 2 model
- **BC.4**: Four-parameter Brain-Cousens hormesis model
- **LL.5**: Five-parameter log-logistic model (asymmetric)
- **EXD.3**: Three-parameter exponential decay model

```{r model-selection}
# Use mselect() to compare multiple models
# This fits each model and compares using AIC
model_comparison <- suppressWarnings(
 mselect(
   ryegrass.LL4,
   fctList = list(LN.4(), W1.4(), W2.4(), BC.4(), LL.5(), EXD.3())
   )
 )
model_comparison
```

**Understanding the Output:**

The table shows:
- **logLik**: Log-likelihood (higher is better, but penalized for parameters)
- **IC**: Information criterion (AIC by default; **lower is better**)
- **Res var**: Residual variance (lower is better)
- **Lack of fit**: P-value for lack-of-fit test (non-significant is better)

Models are sorted by IC (AIC), with the best-fitting model at the top.

### Selecting the Best Model

```{r best-model}
# Based on mselect results, fit the best model
# (In this example, we'll use the model with lowest AIC from the comparison)
# For ryegrass data, typically W1.4 or LL.4 performs well

ryegrass.best <- drm(rootl ~ conc, data = ryegrass, fct = W1.4())

# Summary of best model
summary(ryegrass.best)

# ED estimates for best model
ed_best <- ED(ryegrass.best, respLev = c(10, 20, 50), interval = "delta")
ed_best
```

### Visual Comparison of Models

Plotting multiple models together helps visualize differences in fit:

```{r model-comparison-plot, fig.alt="Comparison of LL.4 and W1.4 model fits showing two overlapping dose-response curves in blue (LL.4) and red dashed line (W1.4)"}
# Plot initial LL.4 model
plot(ryegrass.LL4, type = "all",
     main = "Comparison: LL.4 vs W1.4 Models",
     xlab = "Ferulic acid concentration (mM)",
     ylab = "Root length (cm)",
     lwd = 2, cex = 1.2, col = "blue",
     legend = FALSE)

# Overlay the best model (W1.4)
plot(ryegrass.best, add = TRUE, type = "none", lwd = 2, col = "red", lty = 2)

# Add legend
legend("topright", legend = c("LL.4 (initial)", "W1.4 (best)"),
       col = c("blue", "red"), lwd = 2, lty = c(1, 2), cex = 1.1)
```

### Comparing ED Estimates Between Models

```{r ed-comparison-models}
# Compare EC50 estimates between models
cat("EC50 from LL.4 model:\n")
ED(ryegrass.LL4, respLev = 50, interval = "delta")

cat("\nEC50 from W1.4 model:\n")
ED(ryegrass.best, respLev = 50, interval = "delta")
```

**Important Notes:**
- Different models may yield different ED estimates
- Model selection should be based on both statistical criteria (AIC) and biological plausibility
- Small differences in AIC (< 2) suggest models are essentially equivalent

## Step 5: Model-Averaged ED Estimation

When multiple models fit similarly well, model averaging provides a robust approach that accounts for model uncertainty. The `maED()` function computes model-averaged ED estimates using AIC-based weights.

### Computing Model-Averaged EDs

```{r model-averaging}
# Model-averaged EC50 estimation using top 3 models
# Based on our mselect results, we'll average over several competitive models
ma_results <- maED(ryegrass.LL4,
                   fctList = list(W1.4(), W2.4(), LL.5()),
                   respLev = 50,
                   interval = "buckland")

ma_results
```

**Understanding Model Averaging:**

- Each model receives a weight based on its AIC value
- Better-fitting models (lower AIC) receive higher weights
- The final estimate is a weighted average across models
- Confidence intervals account for both parameter uncertainty and model uncertainty

### Comparing Single-Model vs Model-Averaged Estimates

```{r ma-comparison}
# Compare model-averaged EC50 with single-model estimates
cat("Single model (W1.4) EC50:\n")
ED(ryegrass.best, respLev = 50, interval = "delta")

cat("\nModel-averaged EC50 (top 3 models):\n")
print(ma_results)
```

**When to Use Model Averaging:**
- Multiple models have similar AIC values (ΔAIC < 2-4)
- You want robust estimates that don't depend on selecting a single model
- Regulatory or risk assessment contexts requiring conservative estimates

**When to Use Single Model:**
- One model is clearly superior (ΔAIC > 10)
- Strong biological rationale for a specific model form
- Simpler interpretation needed

## Step 6: Impact of Fixing Asymptotes

The upper and lower asymptotes (parameters `d` and `c`) can be estimated from the data or fixed based on prior knowledge. Understanding when and how to fix these parameters is crucial for proper model fitting.

### Understanding Asymptote Parameters

- **d (upper limit)**: Response at zero dose (control response)
- **c (lower limit)**: Response at infinite dose (maximal effect)

### Models with Different Asymptote Constraints

```{r asymptote-comparison}
# LL.4: Both asymptotes free (4 parameters)
ryegrass.LL4 <- drm(rootl ~ conc, data = ryegrass, fct = LL.4())

# LL.3: Lower asymptote fixed at 0 (3 parameters)
ryegrass.LL3 <- drm(rootl ~ conc, data = ryegrass, fct = LL.3())

# LL.3u: Upper asymptote fixed at 1 (3 parameters)
# Note: Requires normalized data for this to be meaningful
ryegrass_norm <- ryegrass
ryegrass_norm$rootl_norm <- ryegrass$rootl / max(ryegrass$rootl)
ryegrass.LL3u <- drm(rootl_norm ~ conc, data = ryegrass_norm, fct = LL.3u())

# LL.2: Both asymptotes fixed (2 parameters)
ryegrass.LL2 <- drm(rootl_norm ~ conc, data = ryegrass_norm, fct = LL.2())

# Compare models
cat("LL.4 (both free):\n")
summary(ryegrass.LL4)

cat("\nLL.3 (lower = 0):\n")
summary(ryegrass.LL3)

cat("\nAIC Comparison:\n")
cat("LL.4 (4 params):", AIC(ryegrass.LL4), "\n")
cat("LL.3 (3 params):", AIC(ryegrass.LL3), "\n")
```

### Visual Comparison of Constrained Models

```{r asymptote-plot, fig.alt="Comparison of LL.4 and LL.3 models showing effect of asymptote constraints with black solid line (LL.4) and blue dashed line (LL.3)"}
# Plot models with different constraints
plot(ryegrass.LL4, type = "all",
     main = "Effect of Asymptote Constraints",
     xlab = "Ferulic acid concentration (mM)",
     ylab = "Root length (cm)",
     lwd = 2, col = "black", legend = FALSE, cex = 1.2)

plot(ryegrass.LL3, add = TRUE, type = "none", lwd = 2, col = "blue", lty = 2)

legend("topright",
       legend = c("LL.4 (both free)", "LL.3 (c = 0)"),
       col = c("black", "blue"),
       lwd = 2, lty = c(1, 2),
       cex = 1.1)
```

### Implications of Fixing Asymptotes

**Benefits of Fixing Asymptotes:**
1. **Reduced parameter count**: Simpler model, fewer parameters to estimate
2. **Improved stability**: Fewer parameters can mean more stable fits
3. **Biological relevance**: Incorporating prior knowledge (e.g., c = 0 when complete inhibition is impossible)
4. **Identifiability**: Some datasets may not contain enough information to estimate all parameters

**When to Fix Asymptotes:**
- **Fix c = 0** when:
  - Response cannot go below zero (e.g., growth, survival)
  - Biological knowledge indicates complete inhibition doesn't occur
  - Data doesn't extend to high enough doses to estimate c

- **Fix d** when:
  - Control response is known from independent measurements
  - Data is normalized to a known maximum (e.g., 100%)
  - You want to focus on relative potency comparisons

**When to Keep Asymptotes Free:**
- Data extends over a wide dose range
- Both asymptotes are clearly identifiable in the data
- No strong prior knowledge about asymptote values
- Model comparison/selection workflow

### Effect on ED Estimates

```{r asymptote-ed}
# Compare ED estimates with different constraints
cat("EC50 with LL.4 (both asymptotes free):\n")
ED(ryegrass.LL4, respLev = 50, interval = "delta")

cat("\nEC50 with LL.3 (lower asymptote = 0):\n")
ED(ryegrass.LL3, respLev = 50, interval = "delta")
```

**Important Note:** The choice of asymptote constraints can substantially affect ED estimates, especially for EC10 and EC20 values which depend more heavily on the asymptote values than EC50.

## Step 7: Overview of Available Models

The `drc` package provides numerous dose-response models suitable for different types of data and biological mechanisms. Understanding which model to use is crucial for proper analysis.

### Monotonic (Non-Hormesis) Models

Monotonic models describe dose-response relationships that are either strictly increasing or strictly decreasing. These are appropriate when the response changes consistently in one direction as dose increases.

#### Log-Logistic Models (LL family)

**Characteristics:**
- Symmetric on log-dose scale
- Most commonly used in toxicology
- S-shaped curve
- Parameters: b (slope), c (lower), d (upper), e (ED50)

**Variants:**
- `LL.2()`: 2 parameters (c=0, d=1 fixed)
- `LL.3()`: 3 parameters (c=0)
- `LL.3u()`: 3 parameters (d=1)
- `LL.4()`: 4 parameters (most flexible)
- `LL.5()`: 5 parameters (asymmetric, f parameter)

**Best for:**
- General dose-response data
- Toxicity studies
- EC50/ED50 estimation

**Example:**
```{r ll-example, fig.alt="Log-logistic model (LL.4) fitted to ryegrass data showing typical S-shaped dose-response curve"}
# Standard application of log-logistic model
example.LL <- drm(rootl ~ conc, data = ryegrass, fct = LL.4())
plot(example.LL, main = "Log-Logistic Model (LL.4)")
```

#### Weibull Models (W1 and W2 families)

**Characteristics:**
- Asymmetric on log-dose scale
- Two types: W1 (increasing asymmetry) and W2 (decreasing asymmetry)
- Flexible shape
- Same parameter structure as log-logistic

**Variants:**
- `W1.2()`, `W1.3()`, `W1.4()`: Weibull type 1
- `W2.2()`, `W2.3()`, `W2.4()`: Weibull type 2

**Best for:**
- Data with asymmetric dose-response curves
- Time-to-event data
- Germination/mortality studies

**Example:**
```{r weibull-example, fig.alt="Comparison of Weibull Type 1 (blue) and Type 2 (red dashed) models showing asymmetric dose-response curves"}
# Weibull models often fit plant growth data well
example.W1 <- drm(rootl ~ conc, data = ryegrass, fct = W1.4())
example.W2 <- drm(rootl ~ conc, data = ryegrass, fct = W2.4())

# Compare
plot(example.W1, type = "all", main = "Weibull Type 1 vs Type 2",
     lwd = 2, col = "blue", legend = FALSE)
plot(example.W2, add = TRUE, type = "none", lwd = 2, col = "red", lty = 2)
legend("topright", legend = c("W1.4", "W2.4"),
       col = c("blue", "red"), lwd = 2, lty = c(1, 2))
```

#### Log-Normal Models (LN family)

**Characteristics:**
- Based on log-normal distribution
- Symmetric on log-dose scale
- Similar to log-logistic but different tail behavior

**Variants:**
- `LN.2()`, `LN.3()`, `LN.3u()`, `LN.4()`

**Best for:**
- Data with normal distribution on log scale
- Particle size distributions
- Alternative to log-logistic when AIC suggests

**Example:**
```{r lognormal-example}
example.LN <- drm(rootl ~ conc, data = ryegrass, fct = LN.4())
```

#### Exponential Decay Models (EXD family)

**Characteristics:**
- Exponential decrease
- No lower asymptote (unless constrained)
- Simpler than sigmoidal models

**Variants:**
- `EXD.2()`: 2 parameters
- `EXD.3()`: 3 parameters

**Best for:**
- Exponential decay processes
- Radioactive decay
- Simple inhibition without clear asymptote

**Example:**
```{r exd-example}
example.EXD <- drm(rootl ~ conc, data = ryegrass, fct = EXD.3())
```

### Hormesis (Non-Monotonic) Models

Hormesis describes a biphasic dose-response relationship where low doses stimulate a response (increase) while high doses inhibit (decrease). This creates a characteristic inverted U-shape or J-shape curve.

#### Brain-Cousens Models (BC family)

**Characteristics:**
- Adds hormesis parameter to log-logistic model
- Peak response at intermediate dose
- Widely used for hormetic data

**Variants:**
- `BC.4()`: 4 parameters (c=0 fixed)
- `BC.5()`: 5 parameters (all free)

**Parameters:**
- Standard LL parameters plus `f`: hormesis parameter (controls magnitude of stimulation)

**Best for:**
- Plant growth stimulation at low herbicide doses
- Pharmaceutical hormesis
- Toxicological hormesis

**Example (conceptual):**
```{r bc-example, eval=FALSE}
# Example with hormetic data (not ryegrass which is monotonic)
# hormetic.model <- drm(response ~ dose, data = hormetic_data, fct = BC.5())
# plot(hormetic.model)
```

#### Cedergreen-Ritz-Streibig Models (CRS family)

**Characteristics:**
- More flexible hormesis models
- Multiple parameterizations (a, b, c variants)
- Better for pronounced hormesis

**Variants:**
- `CRS.4a()`, `CRS.4b()`, `CRS.4c()`: 4-parameter variants
- `CRS.5a()`, `CRS.5b()`, `CRS.5c()`: 5-parameter variants
- `CRS.6()`: 6 parameters (most flexible)

**Best for:**
- Strong hormesis effects
- When BC models don't fit well
- Detailed hormesis characterization

#### U-Shaped Cedergreen Models (UCRS family)

**Characteristics:**
- U-shaped response (opposite of hormesis)
- Low and high doses harmful, intermediate doses beneficial
- Less common than hormesis

**Variants:**
- `UCRS.4a()`, `UCRS.4b()`, `UCRS.4c()`
- `UCRS.5a()`, `UCRS.5b()`, `UCRS.5c()`

**Best for:**
- Essential nutrients (deficiency and toxicity)
- Biphasic therapeutic responses

### Model Selection Decision Tree

```
Is your dose-response curve monotonic?
│
├─ YES (Monotonic/No Hormesis)
│  │
│  ├─ Standard S-shaped curve? → Start with LL.4
│  ├─ Asymmetric curve? → Try W1.4 or W2.4
│  ├─ Simple decay? → Try EXD.3
│  └─ Unknown? → Compare LL.4, W1.4, W2.4, LN.4 using mselect()
│
└─ NO (Non-Monotonic/Hormesis)
   │
   ├─ Inverted U-shape (stimulation then inhibition)? → Try BC.5
   ├─ Strong hormesis? → Try CRS.5a
   ├─ U-shaped (harm-benefit-harm)? → Try UCRS.5a
   └─ Unknown? → Compare BC.5, CRS.5a with mselect()
```

### Practical Recommendations

1. **Start Simple**: Begin with LL.4 or W1.4 for monotonic data
2. **Use Model Selection**: Always compare multiple models with `mselect()`
3. **Check Residuals**: Visual diagnostics are essential
4. **Consider Biology**: Model choice should make biological sense
5. **Parameter Constraints**: Use simpler models (LL.3, LL.2) when appropriate
6. **Hormesis Testing**: If you suspect hormesis, explicitly test with BC or CRS models

### Comprehensive Model Comparison Example

```{r comprehensive-comparison}
# Compare a wide range of monotonic models for ryegrass data
comprehensive <- suppressWarnings(
   mselect(ryegrass.LL4, nested = TRUE,
                        fctList = list(LL.3(), LL.5(),
                                      W1.3(), W1.4(),
                                      W2.3(), W2.4(),
                                      LN.3(), LN.4(),
                                      EXD.3())) 
)
comprehensive
```

## Conclusion

This vignette has demonstrated a comprehensive workflow for dose-response analysis using the `drc` package. By following these steps, you can:

### Key Takeaways

1. **Always Start with Exploration**: Visualize your data before fitting models
2. **Fit Multiple Models**: Don't rely on a single model without comparison
3. **Use Visual Diagnostics**: Q-Q plots and residual plots are essential
4. **Perform Statistical Tests**: Use `noEffect()` and `modelFit()` to validate your model
5. **Compare Systematically**: Use `mselect()` with AIC for objective model selection
6. **Consider Model Averaging**: Use `maED()` when multiple models fit similarly
7. **Understand Parameter Constraints**: Know when to fix or free asymptotes (c, d parameters)
8. **Choose Models Based on Data Type**: Distinguish between monotonic and hormetic responses

### Common Pitfalls to Avoid

1. **Fitting only one model**: Always compare alternatives
2. **Ignoring diagnostics**: Visual and statistical checks are crucial
3. **Over-parameterization**: More parameters isn't always better
4. **Inappropriate constraints**: Don't fix parameters without justification
5. **Ignoring biology**: Statistical fit should align with biological plausibility
6. **Using hormesis models for monotonic data**: This can lead to spurious hormesis
7. **Not reporting confidence intervals**: Point estimates without uncertainty are incomplete

### Recommended Workflow Summary

1. **Explore** your data with plots
2. **Fit** an initial general model (e.g., LL.4)
3. **Assess** fit visually (Q-Q plots, residual plots)
4. **Test** statistically (noEffect, modelFit)
5. **Compare** multiple models (mselect)
6. **Select** the best model or use model averaging
7. **Estimate** EDs/ECs with appropriate confidence intervals
8. **Evaluate** parameter constraints if needed
9. **Interpret** results in biological context
10. **Report** model choice, fit statistics, and ED estimates with CIs

### Further Resources

- See `?drm` for detailed function documentation
- See `?LL.4`, `?W1.4`, etc. for specific model documentation
- See `?mselect` for model selection details
- See `?ED` for effective dose estimation options
- See the "Understanding NEC Models" vignette for threshold models

## References

Ritz, C., Baty, F., Streibig, J. C., Gerhard, D. (2015). Dose-Response Analysis Using R. *PLOS ONE*, **10**(12), e0146021.

Ritz, C., Streibig, J. C. (2005). Bioassay analysis using R. *Journal of Statistical Software*, **12**(5), 1-22.

Brain, P., Cousens, R. (1989). An equation to describe dose-responses where there is stimulation of growth at low doses. *Weed Research*, **29**, 93-96.

Cedergreen, N., Ritz, C., Streibig, J. C. (2005). Improved empirical models describing hormesis. *Environmental Toxicology and Chemistry*, **24**, 3166-3172.

Inderjit, Streibig, J. C., Olofsdotter, M. (2002). Joint action of phenolic acid mixtures and its significance in allelopathy research. *Physiologia Plantarum*, **114**, 422-428.

## See Also

- `vignette("nec-models")` - Understanding NEC Models in the drc Package
- `?drm` - Main function for fitting dose-response models
- `?ED` - Estimating effective doses
- `?mselect` - Model selection
- `?modelFit` - Goodness-of-fit testing
- `?plot.drc` - Plotting dose-response curves
